---
title: "Comparing totalVI and OSCA book CITE-seq analyses"
author: "OSCA book authors and Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Comparing totalVI and OSCA book CITE-seq analyses}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Overview

This vignette endeavors to put Bioconductor and scvi-tools
together to help understand how different data structures and methods
relevant to CITE-seq analysis contribute to interpretation
of CITE-seq exeperiments.

The scvi-tools tutorial (for version 0.20.0)
analyzes a pair of 10x PBMC CITE-seq experiments (5k and 10k cells).
Chapter 12 of the OSCA book analyzes only the 10k dataset.

# Technical steps to facilitate comparison

## Acquire software
```{r getsoft, message=FALSE}
library(SingleCellExperiment)
library(scater)
library(scviR)
```

## Obtain key data representations

```{r getdat, message=FALSE}
ch12sce = get_ch12sce()
ch12sce
```

```{r getdat2, message=FALSE}
fullvi = get_totalVI_5k10k_adata()
fullvi
```

## Assemble a SingleCellExperiment with totalVI outputs

### Acquire cell identities and batch labels

```{r basicvecs}
totvi_cellids = rownames(fullvi$obs)
totvi_batch = fullvi$obs$batch
```

### Acquire quantifications and latent space positions

```{r latent}
totvi_latent = fullvi$obsm$get("X_totalVI")
totvi_umap = fullvi$obsm$get("X_umap")
totvi_denoised_rna = fullvi$layers$get("denoised_rna")
totvi_denoised_protein = fullvi$obsm$get("denoised_protein")
totvi_leiden = fullvi$obs$leiden_totalVI
```


### Drop 5k data from all

```{r remove5k}
is5k = which(totvi_batch == "PBMC5k")
totvi_cellids = totvi_cellids[-is5k]
totvi_latent = totvi_latent[-is5k,]
totvi_umap = totvi_umap[-is5k,]
totvi_denoised_rna = totvi_denoised_rna[-is5k,]
totvi_denoised_protein = totvi_denoised_protein[-is5k,]
totvi_leiden = totvi_leiden[-is5k]
```

### Label the rows of components

```{r rown}
rownames(totvi_latent) = totvi_cellids
rownames(totvi_umap) = totvi_cellids
rownames(totvi_denoised_rna) = totvi_cellids
rownames(totvi_denoised_protein) = totvi_cellids
names(totvi_leiden) = totvi_cellids
```

### Find common cell ids

In this section we reduce the cell collections
to cells common to the chapter 12 and totalVI
datasets.

```{r findcomm}
comm = intersect(totvi_cellids, ch12sce$Barcode)
```

### Build the totalVI SingleCellExperiment

```{r dosce}
# select and order
totvi_latent = totvi_latent[comm,]
totvi_umap = totvi_umap[comm,]
totvi_denoised_rna = totvi_denoised_rna[comm,]
totvi_denoised_protein = totvi_denoised_protein[comm,]
totvi_leiden = totvi_leiden[comm]
 
# organize the totalVI into SCE with altExp

totsce = SingleCellExperiment(SimpleList(logcounts=t(totvi_denoised_rna))) # FALSE name
nn = SingleCellExperiment(SimpleList(logcounts=t(totvi_denoised_protein))) # FALSE name
reducedDims(nn) = list(UMAP=totvi_umap)
altExp(totsce) = nn
altExpNames(totsce) = "denoised_protein"
totsce$leiden = totvi_leiden
altExp(totsce)$leiden = totvi_leiden
altExp(totsce)$ch12.clusters = altExp(ch12sce[,comm])$label
```

### Reduce the chapter 12 dataset to the cells held in common

```{r trim12}
colnames(ch12sce) = ch12sce$Barcode
ch12sce_matched = ch12sce[, comm]
```

# Key outputs of the chapter 12 analysis

## Clustering and projection based on the ADT quantifications

The TSNE projection of the normalized ADT quantifications and
the [walktrap](https://arxiv.org/abs/physics/0512106) cluster assignments
are produced for the cells common to the two approaches.

```{r lkkey}
plotTSNE(altExp(ch12sce_matched), color_by="label", text_by="label")
```
 
## Cluster profiles based on averaging ADT quantities within clusters

This heatmap uses precomputed cluster averages that are
lodged in the metadata element of the SingleCellExperiment.
Colors represent the log2-fold change from the grand average across all clusters.
```{r lkadt}
adt_profiles(ch12sce_matched)
```

## Marker expression patterns in mRNA-based sub-clusters of ADT-based clusters

We enhance the annotation of the list of subclusters retrieved
using `get_ch12_allsce` and then drill into mRNA-based
subclusters of ADT-based cluster 3 to compare expression levels
of three genes.
```{r dosubcl}
ch12_allsce = get_ch12_allsce() 
ch12_allsce = lapply(ch12_allsce, function(x) {
   colnames(x)= x$Barcode; 
   cn = colnames(x); 
   x = x[,intersect(cn,comm)]; x})
of.interest <- "3"
markers <- c("GZMH", "IL7R", "KLRB1")
plotExpression(ch12_allsce[[of.interest]], x="subcluster",
    features=markers, swap_rownames="Symbol", ncol=3)
```
There is a suggestion of a boolean basis for subcluster
identity, depending on low or high expression of the selected genes.

## Graduated relationships between mRNA and surface protein expression

Following the exploration in OSCA chapter 12, cluster 3 is analyzed
for a regression association between expression
measures of two genes and the ADT-based abundance of CD127.
```{r lkgra}
plotExpression(ch12_allsce[["3"]], x="CD127", show_smooth=TRUE, show_se=FALSE, 
    features=c("TPT1", "GZMH"), swap_rownames="Symbol")
```
