---
title: "scvi-tools CITE-seq tutorial in R, using serialized tutorial components"
author: "Follows scvi-tools doc, Gayoso, Steier et al. DOI 10.1038/s41592-020-01050-x"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{scvi-tools CITE-seq tutorial in R, using serialized tutorial components}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Overview

[scvi-tools](https://scvi-tools.org/) is
an element of the [scverse](https://scverse.org/) toolchest for
single-cell omics data analysis.

The scviR package is a very elementary approach to interfacing
between R, Bioconductor and scvi-tools.  

- scvi-tools components develop rapidly; we are
using [basilisk](https://bioconductor.org/packages/basilisk)
to manage R/python interoperation, and as of current
writing we work with version 0.20.0 of scvi-tools.  Specific
versions of python components are enumerated in the file R/basilisk.R.
- A docker container based on a PyTorch-oriented image in
the [NVIDIA container registry](nvcr.io/nvidia/pytorch:22.12-py3) includes
R 4.2.2 and sufficient python 3 infrastructure to use scvi-tools 0.18.0
and scviR 0.0.1.  A Dockerfile is in scviR/inst/Docker;
`vjcitn/nvidpt_bioc:0.0.2` can be pulled from dockerhub, it reports
size 26.5GB.  *This will need to be updated to deal with scvi-tools 0.20.0.*
- Code presented here follows [the colab vignette
for scvi-tools 0.18.0](https://colab.research.google.com/github/scverse/scvi-tutorials/blob/0.18.0/totalVI.ipynb).  *We will check for modifications in the scvi-tools 0.20.0 notebook*.


# A CITE-seq example

## Retrieval of PBMC data
The following chunk will
acquire (and cache, using BiocFileCache) a preprocessed version of the 10k and 5k combined
CITE-seq experiments from the scvi-tools data repository.


```{r doinit}
library(scviR)
adref = get_citeseq_5k10k_pbmcs()
adref
```

## Retrieval of fitted VAE

The totalVI variational autoencoder was fit
with these data.  A fitted version is retrieved and cached
using

```{r dogetv}
vae = get_citeseq_tutvae()
vae
```

The structure of the VAE is reported using
```{r lkmod, eval=FALSE}
vae$module
```
This is quite voluminous and is provided in an appendix.

## Trace of (negative) ELBO values

```{r lkelb}
h = vae$history
npts = nrow(h$elbo_train)
plot(seq_len(npts), as.numeric(h$elbo_train[[1]]), ylim=c(1200,1400), 
  type="l", col="blue", main="Negative ELBO over training epochs")
graphics::lines(seq_len(npts), as.numeric(h$elbo_validation[[1]]), type="l", col="orange")
```

## Normalized quantities

On a CPU, the following can take a long time.
```{r getn, eval=FALSE}
NE = vae$get_normalized_expression(n_samples=25L, 
    return_mean=TRUE, 
    transform_batch=c("PBMC10k", "PBMC5k")
)
```
We provide the totalVI-based denoised quantities in
```{r getdenoise}
denoised = get_totalVI_normalized_5k10k()
vapply(denoised, dim, integer(2))
```
Note that these have features as columns, samples (cells)
as rows.
```{r lkn}
utils::head(colnames(denoised$rna_nmlzd))
utils::head(colnames(denoised$prot_nmlzd))
```

## UMAP projection of Leiden clustering in the totalVI latent space

We have stored a fully loaded anndata instance for retrieval
to inspect the latent space and clustering produced
by the tutorial notebook procedure.  The images produced here
do not agree exactly with what I see in the colab pages for
0.18.0 and 0.19.0.  The process was run in Jetstream2, not in colab.

```{r getproj, fig.height=6}
full = get_totalVI_5k10k_adata()
# class distribution
cllabs = full$obs$leiden_totalVI
blabs = full$obs$batch
table(cllabs)
um = full$obsm$get("X_umap") 
dd = data.frame(umap1=um[,1], umap2=um[,2], clust=factor(cllabs), batch=blabs)
library(ggplot2)
ggplot(dd, aes(x=umap1, y=umap2, colour=clust)) + geom_point(size=.05)
```

Effectiveness at accommodating the two-batch design is suggested by the
mixed representation of the batches in all the Leiden clusters.
```{r getba}
ggplot(dd, aes(x=umap1, y=umap2, colour=factor(batch))) + geom_point(size=.05)
```

## Protein abundances in projected clusters

We focus on four of the ADT.

```{r lknn,fig.width=8}
pro4 = denoised$prot_nmlzd[,1:4]
names(pro4) = gsub("_.*", "", names(pro4))
library(reshape2)
wprot = cbind(dd, pro4)
mm = melt(wprot, id.vars=c("clust", "batch", "umap1", "umap2"))
utils::head(mm,3)
ggplot(mm, aes(x=umap1, y=umap2, colour=log1p(value))) + 
   geom_point(size=.1) + facet_grid(.~variable)
```

# Appendix: The VAE module

The structure of the VAE is reported using
```{r lkmod2, eval=TRUE}
vae$module
```
